$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines

### 2020-03-28 write_up/ms.tex
redraft of article

The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \citep{lambert2018Student}. MCMC methods are thus created because of an asymptotic property: that given an infinite number of samples, their sampling distribution approaches the posterior (under general conditions). In the face of only these asymptotic guarantees, it is actually quite miraculous that MCMC samplers \textit{can} often practically converge to the posterior in only a (relatively small) finite number of samples.

@@

Here, we introduce $R^*$, a new convergence metric. This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a sample the chain that generated it. Rephrased, it should not be possible to predict the chain that \textit{caused} a sample. In this vein, we use a supervised machine learning (ML) approach to measure convergence. Specifically, we train a ML model to classify the chain that generated each observation. By evaluating the performance of the model on a held-out test set, this provides a new convergence metric. To maximise predictive accuracy, ML models naturally exploit differences between the full joint distributions of each chain, which means they are sensitive to variations across the joint distribution of target model dimensions unlike most existent approaches. This statistic, unlike its $\hat{R}$ cousins, is univariate: one model provides a single $R^*$, whereas $\hat{R}$ has separate values for each parameter. However, the ML models we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy. For our machine learning model, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}. For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\hat{R}$ measures calculated (typically less than 1 second to train then test). It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies. In addition, since GBMs can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence. $R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.

@@

These two cases, whilst simple, illustrate the basis of our approach. To determine if a set of Markov chains have converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each sample belongs. By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains. By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\hat{R}$ \citep{gelman2013bayesian}. In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence. Algorithm \ref{alg:R_star} gives an explicit recipe for calculating $R^*$.

@@

The ML model we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \citep{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours. Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' package \citep{greenwell2019package}. The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters. We split each chain's samples into randomly sampled training and testing tests: here, we use 70\% of samples for training and 30\% for testing. The GBM model we use was found to be rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for both these steps for models with 1000 samples), and its predictive performance was insensitive to its hyperparameters. In all the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown (see \url{https://github.com/ben18785/ml-mcmc-convergence} to replicate this functionality).

@@

\section{Results}\label{sec:results}
To illustrate the versatility of $R^*$, we use examples from \citep{vehtari2019rank} and two additional examples based on a multivariate normal model.

@@

To illustrate the consistency of $R^*$, we perform 1000 replicates where, in each case, we generate four $\{X_t\}$ series as described (i.e. with one chain with a lower variance). We then fit a gradient-boosted model to a labelled training set of 2800 randomly chosen samples. The fitted model is then used to classify samples in an independent test set (comprising the remaining 1200 observations) according to the chain which generated them. For each replicate, we then calculate $R^*$ as described in Algorithm \ref{alg:R_star}. In Fig. \ref{fig:ar1}B, we show that the accuracy ratio is well above 1 for all replicates indicating that the chains have not converged.

@@

Since GBMs return a probability simplex for each sample, which, for each chain, indicates the probability that a sample was generated by it, we can also generate a measure of uncertainty in $R^*$. Algorithm \ref{alg:R_star_uncertainty} describes in detail how this can be done. We demonstrate this idea using two datasets: one generated as described above, where one chain (out of four) has a lower variance than the others (we call this the ``unconverged'' data'); and another, where all chains sample from the same distribution (we call this the ``converged'' data). In Fig. \ref{fig:ar1}C, we show the $R^*$ distributions in each case. For the unconverged data, the distribution has its bulk of mass away from 1, indicating lack of convergence. For the converged data, the distribution is centred on 1, indicating convergence.

@@

First, we use the code provided in \citep{vehtari2019rank} to calculate rank-normalised $\hat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk ESS and tail ESS. In all cases, the various quantities were calculated based on chains split into halves. For both dimensions, the two ESS measures were above 3700 and $\hat{R}<1.001$, indicating no issues with convergence.

Next, we estimate the $R^*$ distribution which is shown in Fig. \ref{fig:bivariate}. The mean of this distribution is 1.18, and all the samples are above 1, indicating that the sampling distribution has not converged. By taking account of all the information in the chains, $R^*$ is able to probe issues in joint distribution convergence which are missed by measures that consider only the marginals. As such, we see a role for $R^*$ as an additional convergence metric.

@@

We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \citep{hoffman2014no}. We assume that the Wishart distribution's degrees of freedom parameter is 250, resulting in a distribution with high correlations between dimensions. We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5). First, we used the ``centered'' parameterisation of this model, which is of the form,
%
\begin{equation}
\boldsymbol{x}\sim \mathcal{N}(\boldsymbol{0},\boldsymbol{A}^{-1}),
\end{equation}
%
where $\boldsymbol{x}\in\mathbb{R}^{250}$. For each set of samples, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which are shown in Fig. \ref{fig:mvt}A. From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution. Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (it's mean is 1.02): in this case, $\hat{R}<1.01$ for all parameters, although 61\% of parameters had an $\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.

@@

In GBMs, it is possible to calculate variable importance (see, for example, \citep{friedman2001greedy} and \citep{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power. We now compare these with the more established metrics $\hat{R}$ and ESS as calculated in Stan. For a GBM fitted to the centered model, we plot in Fig. \ref{fig:mvt}B variable importance (here high values mean a variable is more important) versus $\hat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle). From this plot, there appears a weak positive association between GBM's variable importance and $\hat{R}$, although the ranked correlation is not significant (Spearman's rank correlation: $\rho=0.05, S=2468800, p=0.35$). In Fig.  \ref{fig:mvt}C, we plot variable importance versus ESS, which exhibits a strong non-linear negative association (Spearman's rank correlation: $\rho=-0.54, S=4060003, p<0.01$). Since neither of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that the variable importance measure provides information complementary to $\hat{R}$ and ESS.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_three.pdf}}
	\caption{\textbf{Multivariate normal example.} A shows $R^*$ distributions obtained for two MCMC samples from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''); B shows variable importance versus $\hat{R}$ for each parameter as calculated by Stan; and C shows variable importance versus ESS as calculated by Stan. In plots B and C, horizontal jitter was added to the points. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:mvt}
\end{figure}

\subsection{Infinite variance: Cauchy example}
We next explore how $R^*$ can be used to determine convergence for distributions with infinite variance. Like \citep{vehtari2019rank}, we first use Stan to sample from independent standard Cauchy distributions for each element of a 50-element vector $x$,
%
\begin{equation}
x_j\sim \text{Cauchy}(0, 1),\; \text{for } j=1,...,50,
\end{equation}
%
We call this parameterisation the ``nominal'' version of this model.

@@

In Fig. \ref{fig:cauchy}A, we show the $R^*$ distribution under both parameterisations. As shown in \citep{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, samples still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$. The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$. Interesting, however, in contrast to the measures presented in \citep{vehtari2019rank}, the $R^*$ distribution for the alternative parameterisation has a mean above 1, indicating that more MCMC iterations are needed: in Fig. \ref{fig:cauchy}B, we show the result for running each Stan model for ten times as long (and thinning by a factor of 5). In this figure, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has occurred. Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.

@@

To illustrate that $R^*$ provides a reliable metric for capturing convergence, we now calculate a quantitative measure that captures how closely a sampling distribution matches the target. One measure of distributional "closeness" is the KL-divergence, which, in this case, could measure the divergence from the target to the sampling distribution: if the target distribution is known, fitting a kernel density estimator (KDE) to samples allows an approximate (typically univariate) measure of KL-divergence to be calculated. The trouble is, for distributions like the Cauchy with fat tails, fitting a KDE to the samples provides a noisy measure of the sampling distribution in the tails. This means that approximate KL-divergence is unreliable for these types of model. Instead, we choose a measure of distributional discrepancy based around similarity between target quantiles and sample-estimated equivalents. Specifically, we calculate the $R^2$ for the regression of actual quantile values on sample-estimated quantiles, where, if $R^2\sim 1$, the sampling distribution recapitulates well the target quantities. In our example, we consider all percentiles: 0.1\%, 0.2\%,...,99.8\%, 99.9\% and calculate the mean $R^2$ across all 50 dimensions.

@@

Comparing our ``omniscient'' measure of convergence shown in Fig. \ref{fig:cauchy_convergence}A, with the various derived measures, it is clear that all show a similar pattern: as sample size increases, the various statistics tend towards convergence. The rate at which these converge differs though, and $R^*$ (Fig. \ref{fig:cauchy_convergence}D) appears at least, qualitatively, most similar to our ``omniscient'' measure.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy_convergence.pdf}}
	\caption{\textbf{Measuring convergence with MCMC sample size for Cauchy model.} A shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; B shows the maximum value of split-$\hat{R}$ across each of the 50 dimensions of the target; C shows the multivariate split $\hat{R}$ value; D shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}; and E and F show tail- and bulk-ESS. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:cauchy_convergence}
\end{figure}

\subsection{Non-stationary marginals}
To determine whether $R^*$ can detect non-stationary sampling distributions, we recapitulate an example from appendix A in \cite{vehtari2019rank}. This example showed that split-$\hat{R}$ could detect non-convergence caused by shifts in sampling distributions over time: in their case, they analysed chains with common linear trends in mean. Specifically, they first generated four chains by random sampling from a univariate normal distribution, then added a common time trend to each chain, resulting in a univariate distribution whose mean increased during sampling. We first repeat this analysis but using $R^*$ rather than $\hat{R}$: in Fig. \ref{fig:trends_all_dim}, we show these results. In column A, we show the results for $R^*$ calculated on the four chains that were run; column B shows the same calculation but after the chains are split into two equal halves. The different rows show the different sample sizes investigated: 250, 1000 and 4000; the horizontal axis shows the magnitude of time trend added to each sample; for all parameter sets, we run 10 replicates. This plot mirrors Fig. 18 in \cite{vehtari2019rank} and shows that, without splitting the chains, $R^*$ does not increase with trend whereas, after splitting, it does. As expected, split-$R^*$ is more reliably able to detect non-convergence as sample size increases.

These results are what would intuitively be expected: without systematic between-chain variation, it is not possible to reliably determine which of them caused a particular observation. In this case, because all chains exhibited the same secular trends over time, there would not be differences in their marginals. By splitting chains into two -- the first half being the early phase and, the second half being the later phase with higher mean -- this forces differences in the marginals. This meant it was possible to reliably pick whether an observation was caused by an early phase chain or a later one. As such, we recommend that $R^*$ always be calculated using split chains as is recommended for $\hat{R}$ \cite{carpenter2017stan,vehtari2019rank}.

@@

We next consider whether split-$R^*$ can detect non-convergence when only a single dimension trends. In Fig. \ref{fig:trends_one_dim}, we show the effect on $R^*$ of trends of different magnitudes across a range of target dimensions. In the simulations here, all dimensions bar one are stationary; the remaining dimension has a trend added to it. In all cases, split-$R^*$ increased with trend. Indeed, differences in the typical values of this metric were not apparent across the different target dimensions considered. This suggests split-$R^*$ can robustly determine chain identity if only a single dimension has a non-stationary sampling distribution.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_one_dim.pdf}}
	\caption{\textbf{Multivariate trends example with split-$R^*$.} The columns represent different dimensionalities of the target distribution; the rows present different sample sizes. The horizontal axis measures (half) the change in mean across the single dimension that had a trend added to it: a value ``1'' indicates its mean increases by 2 units from the start to end of sampling; all other dimensions (if dimensions exceeded 1) had stationary distributions. At each parameter set, 10 replicates were run and jitter was added to the points. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:trends_one_dim}
\end{figure}

Means that trends over time is one form of non-stationarity; another is a time-varying covariance. Next, we consider a bivariate normal with (constant) standard normal marginals, but where the correlation between dimensions trends over time. Specifically, we allow the correlation to increase from $-\rho$ and $\rho$ throughout the course of simulations and use i.i.d. samples from the process across 4 ``chains''. Again, as before, $R^*$ calculated on unsplit chains is unable to detect this form of non-stationarity, since there are no inter-chain differences in the sampling distribution. Similarly, split-$\hat{R}$ does not detect this form of non-convergence since the marginal distribution across chains does not vary over time (Fig. \ref{fig:trends_joint_distribution}A). By contrast, split-$R^*$ can (Fig. \ref{fig:trends_joint_distribution}B), since it uses all information in the samples, including the covariance structure.

>>>>

The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \citep{lambert2018Student}. MCMC methods are thus created because of an asymptotic property: that given an infinite number of samples, their sampling distribution approaches the posterior (under general conditions). With only these asymptotic guarantees, it is actually quite miraculous that MCMC samplers \textit{can} often approximate the posterior with only a (relatively small) finite number of samples.

@@

Here, we introduce $R^*$, a new convergence metric. This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a sample's value the chain that generated it. Rephrased, it should not be possible to predict the chain that \textit{caused} a sample. In this vein, we use a supervised machine learning (ML) approach to measure convergence. Specifically, we train a ML model to classify the chain that generated each observation. By evaluating the performance of the model on a held-out test set, this provides a new convergence metric. To maximise predictive accuracy, ML models naturally exploit differences between the full joint distributions of each chain, which means they are sensitive to variations across the joint distribution of target model dimensions unlike most existent approaches. This statistic, unlike its $\hat{R}$ cousins, is univariate: one model provides a single $R^*$, whereas $\hat{R}$ has separate values for each parameter. However, the ML models we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy. For our machine learning model, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}. For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\hat{R}$ measures calculated (typically less than 1 second to calculate). It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies. In addition, since GBMs can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence. $R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.

@@

These two cases, whilst simple, illustrate the basis of our approach. To determine if a set of Markov chains has converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each sample belongs. By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains. By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\hat{R}$ \citep{gelman2013bayesian}. In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence. Algorithm \ref{alg:R_star} gives a recipe for calculating $R^*$.

@@

The ML model we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours. Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' package \citep{greenwell2019package}. The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters. We split each chain's samples into randomly sampled training and testing tests: here, we use 70\% of samples for training and 30\% for testing. Our GBM model was rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters. In all the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown (see \url{https://github.com/ben18785/ml-mcmc-convergence} to replicate this functionality).

@@

\section{Results}\label{sec:results}
To illustrate the versatility of $R^*$, we use a range of examples that demonstrate how this statistic fares across a range of scenarios.

@@

To illustrate the consistency of $R^*$, we perform 1000 replicates where, in each case, we generate four $\{X_t\}$ series as described (i.e. where one chain has a lower variance). We then fit a GBM to a labelled training set of 2800 randomly chosen samples. The fitted model is then used to classify samples in an independent test set (comprising the remaining 1200 observations) according to the chain which generated them. For each replicate, we then calculate $R^*$ as described in Algorithm \ref{alg:R_star}. In Fig. \ref{fig:ar1}B, we show that $R^*>1$ for all replicates indicating that the chains have not converged.

@@

Since GBMs return a probability simplex for each sample, which, for each chain, indicates the probability that a sample was generated by it, we can also generate a measure of uncertainty in $R^*$. Algorithm \ref{alg:R_star_uncertainty} describes in detail how this can be done. We demonstrate this idea using two datasets: one generated as described above, where one chain (out of four) has a lower variance than the others (we call this the ``unconverged'' data); and another, where all chains sample from the same distribution (we call this the ``converged'' data). In Fig. \ref{fig:ar1}C, we show the $R^*$ distributions in each case. For the unconverged data, the distribution has its bulk of mass away from 1, indicating lack of convergence. For the converged data, the distribution is centred on 1, indicating convergence.

@@

First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\hat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk ESS and tail ESS. In all cases, the various quantities were calculated based on chains split into halves. For both dimensions, the two ESS measures were above 3700 and $\hat{R}<1.001$, indicating no issues with convergence.

Next, we estimate the $R^*$ distribution using Algorithm \ref{alg:R_star_uncertainty}, which is shown in Fig. \ref{fig:bivariate}. The mean of this distribution is 1.18, and all $R^*$ samples are above 1, indicating that the sampling distribution has not converged. By taking account of all the information in the chains, $R^*$ is able to probe issues in joint distribution convergence which are missed by measures that consider only the marginals.

@@

We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \citep{hoffman2014no}. We assume that the Wishart distribution's degrees of freedom parameter is 250, resulting in a distribution with high correlations between dimensions. We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5). First, we used the ``centered'' parameterisation of this model, which is of the form,
%
\begin{equation}\label{eq:mvt_normal_250}
\boldsymbol{x}\sim \mathcal{N}(\boldsymbol{0},\boldsymbol{A}^{-1}),
\end{equation}
%
where $\boldsymbol{x}\in\mathbb{R}^{250}$. For each set of samples, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which are shown in Fig. \ref{fig:mvt}A. From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution. Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.02): in this case, $\hat{R}<1.01$ for all parameters, although 61\% of parameters had an $\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.

@@

In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power. We now compare these with the more established metrics $\hat{R}$ and ESS as calculated in Stan. For a GBM fitted to the centered model, we plot in Fig. \ref{fig:mvt}B variable importance (here high values mean a variable is more important) versus $\hat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle). From this plot, there appears a weak positive association between GBM's variable importance and $\hat{R}$, although the ranked correlation is not significant (Spearman's rank correlation: $\rho=0.05, S=2468800, p=0.35$). In Fig.  \ref{fig:mvt}C, we plot variable importance versus ESS, which exhibits a strong non-linear negative association (Spearman's rank correlation: $\rho=-0.54, S=4060003, p<0.01$). Since neither of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that variable importance provides information complementary to $\hat{R}$ and ESS.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_three.pdf}}
	\caption{\textbf{Multivariate normal example.} A shows $R^*$ distributions obtained for two MCMC samples from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''); B shows variable importance versus $\hat{R}$ for each parameter as calculated by Stan; and C shows variable importance versus ESS as calculated by Stan. In plots B and C, horizontal jitter was added to the points and a loess fit line with standard errors overlaid. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:mvt}
\end{figure}

\subsection{Wide datasets: multivariate normal}\label{sec:wide}
As the number of parameter dimensions increases, it could be hypthesised that ML algorithms will overfit the data, and, hence, testing set classification would be poor; in being so, this could lead us to falsely conclude that convergence had occurred. To test this hypothesis, we investigated two scenarios based around the multivariate normal. In the first of these, we used the 250-dimensional multivariate normal of eq. \ref{eq:mvt_normal_250} and used 250 post-warm-up iterations (after 250 warm-up iterations) from Stan to calculate $R^*$ distributions as in Algorithm \ref{alg:R_star_uncertainty}. Here, we considered both the centered and non-centered parameterisations, where, in both cases, the number of iterations is comparable to the number of parameters, so the training data is relatively ``wide'': the $R^*$ distribution in each case is shown in Fig. \ref{fig:mvt_wide_both}A. In this figure, it is clear that both $R^*$ distributions are shifted away from 1, indicating non-convergence. The same conclusion is reached if split-$\hat{R}$ is used instead (Fig. \ref{fig:wide_both_diagnostics}A), since, for both parameterisations, there remain some parameters with $\hat{R}>1.01$. Using bulk- or tail-ESS instead, we conclude that the non-centered parameterisation shows signs of convergence whereas the centered does not (Fig. \ref{fig:wide_both_diagnostics}B\&C).

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_wide_both.pdf}}
	\caption{\textbf{Wide data examples: $R^*$ distributions.} A shows the $R^*$ distribution for the 250-dimensional example in \S\ref{sec:wide} with 250 post-warm-up iterations from Stan's NUTS algorithm across both model parameterisations. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:mvt_wide_both}
\end{figure}

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1\textwidth]{../output/wide_both_diagnostics.pdf}}
	\caption{\textbf{Established convergence diagnostics for 250-dimensional normal example.} The top row shows the results for the centered parameterisation; the bottom row for the non-centered. Column A shows split-$\hat{R}$; columns B and C show the bulk- and tail-ESS; in each case the statistics are displayed for all model parameters and were calculated using 250 post-warm-up samples from Stan's NUTS algorithm. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:wide_both_diagnostics}
\end{figure}

We next consider a more challenging example -- a target distribution with 10,000 dimensions. In this case, we assume independent standard normals for each dimension. In Fig. \ref{fig:mvt_wide_both}B, we plot the $R^*$ distribution for two MCMC runs targeting this distributions: one with 400 iterations, the other with 1000. In both cases, the distributions were right of $R^*=1$, indicating non-convergence. These results were also echoed by split-$\hat{R}$, with 28\% of dimensions having $\hat{R}>1.01$ for the 400 iteration case and 3\% for the 1000 iteration case.

Overall, the examples in this section suggest that $R^*$ is robust to wide datasets; we also note that the statistic took comparable time to calculate relative to existing convergence diagnostics on a desktop computer.

\subsection{Infinite variance: Cauchy example}
We next explore how $R^*$ can be used to determine convergence for distributions with infinite variance. Like \cite{vehtari2019rank}, we first use Stan to sample from independent standard Cauchy distributions for each element of a 50-dimensional vector $x$,
%
\begin{equation}
x_j\sim \text{Cauchy}(0, 1),\; \text{for } j=1,...,50.
\end{equation}
%
We call this parameterisation the ``nominal'' version of this model.

@@

In Fig. \ref{fig:cauchy}A, we show the $R^*$ distribution under both parameterisations. As shown in \cite{vehtari2019rank}, the nominal parameterisation results in poor sampling efficiency due to its long tails, meaning that, after 2000 MCMC iterations, samples still contain information about chain identity, and, accordingly, the $R^*$ distribution is shifted rightwards from $R^*=1$. The alternative parameterisation fares better and the $R^*$ distribution is nearer $R^*=1$. Interesting, however, in contrast to the measures presented in \cite{vehtari2019rank}, the $R^*$ distribution for the alternative parameterisation has a mean above 1, indicating that more MCMC iterations are needed: in Fig. \ref{fig:cauchy}B, we show the result for running each Stan model for ten times as long (and thinning by a factor of 5). In this figure, the alternative parameterisation now has an $R^*$ distribution centered on $R^*=1$ and, hence, we are more confident that convergence has occurred. Despite the added iterations, the $R^*$ distribution from the nominal model remains stubbornly away from 1.

@@

To illustrate that $R^*$ provides a reliable metric for capturing convergence, we now calculate a quantitative measure that captures how closely a sampling distribution matches the target. One measure of distributional ``closeness'' is the KL-divergence, which, in this case, could be used to measure the divergence from target to sampling distribution: if the target distribution is known, fitting a kernel density estimator (KDE) to samples allows an approximate (typically univariate) measure of KL-divergence to be calculated for each dimension. The trouble is, for distributions like the Cauchy with fat tails, fitting a KDE to the samples provides a noisy measure of the sampling distribution in the tails. This means that approximate KL-divergence is unreliable for these types of model. Instead, we choose a measure of distributional discrepancy based around similarity between target quantiles and sample-estimated equivalents. Specifically, we calculate the $R^2$ for the regression of actual quantile values on sample-estimated quantiles, where, if $R^2\sim 1$, the sampling distribution recapitulates well the target quantities. In our example, we consider all percentiles: 0.1\%, 0.2\%,...,99.8\%, 99.9\% and calculate the mean $R^2$ across all 50 dimensions.

@@

Comparing our ``omniscient'' measure of convergence shown in Fig. \ref{fig:cauchy_convergence}A, with the various derived measures, all show a similar pattern: as sample size increases, the various statistics tend towards convergence. The rate at which these converge differs though, and $R^*$ (Fig. \ref{fig:cauchy_convergence}D) appears at least, qualitatively, most similar to our ``omniscient'' measure.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/cauchy_convergence.pdf}}
	\caption{\textbf{Measuring convergence for the Cauchy model.} A shows a measure of convergence, the mean quantile $R^2$, that requires knowing the target distribution; B shows the maximum value of split-$\hat{R}$ across each of the 50 dimensions of the target; C shows the multivariate split-$\hat{R}$ value; D shows the value of split-$R^*$ as calculated by Algorithm \ref{alg:R_star}; and E and F show tail- and bulk-ESS. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:cauchy_convergence}
\end{figure}

\subsection{Non-stationary marginals}
To determine whether $R^*$ can detect non-stationary sampling distributions, we recapitulate an example from appendix A in \cite{vehtari2019rank}. This example showed that split-$\hat{R}$ could detect non-convergence caused by shifts in sampling distributions over time: in their case, they analysed chains with common linear trends in mean. Specifically, they first generated four chains by random sampling from a univariate normal distribution, then added a common time trend to each chain, resulting in a univariate distribution whose mean increased during sampling. We first repeat this analysis but using $R^*$ rather than $\hat{R}$: in Fig. \ref{fig:trends_all_dim}, we show these results. In column A, we show the results for $R^*$ calculated on the four chains that ran; column B shows the same calculation but after the chains are split into two equal halves. The rows show the range of sample sizes investigated: 250, 1000 and 4000; the horizontal axis shows the magnitude of time trend added to each sample; for all parameter sets, we run 10 replicates. This plot mirrors Fig. 18 in \cite{vehtari2019rank} and shows that, without splitting the chains, $R^*$ does not increase with trend whereas, after splitting, it does. As expected, split-$R^*$ is more reliably able to detect non-convergence as sample size increases.

These results make intuitive sense: without systematic between-chain variation, it is not possible to reliably determine which of them caused a particular observation. In this case, because all chains exhibited the same secular trends over time, there would not be differences in their marginals. By splitting chains into two -- the first half being the early phase, and the second half being the later phase with higher mean -- this forces differences in the marginals. This meant it was possible to reliably pick whether an observation was caused by an early phase chain or a later one. As such, we recommend that $R^*$ always be calculated using split chains as is recommended for $\hat{R}$ \citep{carpenter2017stan,vehtari2019rank}.

@@

We next consider whether split-$R^*$ can detect non-convergence when only a single dimension trends. In Fig. \ref{fig:trends_one_dim}, we show how $R^*$ performs across a range of target dimensions. In the simulations here, all dimensions bar one are stationary; the remaining dimension has a linear trend added to it. In all cases, split-$R^*$ increased with trend. Indeed, differences in the typical values of this metric were not apparent across the different target dimensions considered. This suggests split-$R^*$ can robustly determine chain identity if only a single dimension has a non-stationary sampling distribution.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_one_dim.pdf}}
	\caption{\textbf{Multivariate trends example with split-$R^*$.} The columns present different dimensionalities of the target distribution; the rows present different sample sizes. The horizontal axis measures (half) the change in mean across the single dimension that had a trend added to it: a value ``1'' indicates its mean increases by 2 units from the start to end of sampling; all other dimensions (if dimensions exceeded 1) had stationary distributions. At each parameter set, 10 replicates were run and jitter was added to the points. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:trends_one_dim}
\end{figure}

Means that trend over time is one form of non-stationarity; another is a time-varying covariance. Next, we consider a bivariate normal with (constant) standard normal marginals, but where the correlation between dimensions trends over time. Specifically, we allow the correlation to increase linearly from $-\rho$ and $\rho$ throughout the course of simulations, and use i.i.d. samples from the process across 4 ``chains''. Again, as before, $R^*$ calculated on unsplit chains is unable to detect this form of non-stationarity, since there are no inter-chain differences in the sampling distribution. Similarly, split-$\hat{R}$ does not detect this form of non-convergence since the marginal distribution across chains does not vary over time (Fig. \ref{fig:trends_joint_distribution}A). By contrast, split-$R^*$ can (Fig. \ref{fig:trends_joint_distribution}B), since it uses all information in the samples, including the covariance structure.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_joint_distribution.pdf}}
	\caption{\textbf{Bivariate normal with trending correlation example.} Column A shows the results for split-$\hat{R}$; column B for $R^*$; the rows present the differing sample sizes. The horizontal axis measures (half) the change in correlation across the whole sample: so a value ``0.5'' indicates the correlation increases by 1 unit (from -0.5 to 0.5) from the start to end of sampling. At each parameter set, 10 replicates were run and jitter was added to the points. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:trends_joint_distribution}
\end{figure}
